# Churn customer analysis of the telekom data"


### Contributions

- Oleksandr Zadorozhnyi



### Data


**Description**

In this notebook we analyze a dataset on the churn rate of telecom operator clients. The dataset contains 50000 rows (customers) and 20 columns (variables/features). The 20 variables are:

```         
state: Categorical, for the 5151 states and the District of Columbia.

area.code: Categorical.

account.length: count, how long account has been active.

voice.plan: Categorical, yes or no, voice mail plan.

voice.messages: Count, number of voice mail messages.

intl.plan: Categorical, yes or no, international plan.

intl.mins: Continuous, minutes customer used service to make international calls.

intl.calls: Count, total number of international calls.

intl.charge: Continuous, total international charge.

day.mins: Continuous, minutes customer used service during the day.

day.calls: Count, total number of calls during the day.

day.charge: Continuous, total charge during the day.

eve.mins: Continuous, minutes customer used service during the evening.

eve.calls: Count, total number of calls during the evening.

eve.charge: Continuous, total charge during the evening.

night.mins: Continuous, minutes customer used service during the night.

night.calls: Count, total number of calls during the night.

night.charge: Continuous, total charge during the night.

customer.calls: Count, number of calls to customer service.

churn: Categorical, yes or no. Indicator of whether the customer has left the company (yes or no).
```


```{r}
#| message: false
df <- readr::read_csv("data/churn.csv")
df
```





***Setup of the problem***

We explore the application of data-driven causal inference, which allows to infer causal relationships purely from observational data. The aim is to identify the reasons which may be the causes for a user of telecom company to become churn and use them to predict the level of churn for a customer while intervening on some of the parameters in the model.



### Load packages

```{r}
#library("liver")
library(qgraph)
library(caret)
library(igraph)
library(bnlearn)
library(tidyverse)
library(tidymodels)
```


### Data exploration

How many loyal and churned clients are in the dataset? Does the distribution of churn vary across the three different areas in the dataset?

```{r}
df |> 
  count(churn)

df |> 
  group_by(area.code) |> 
  count(churn)

```

What is the average number of minutes that churned and loyal clients spoke to the operator in one day?

```{r}
df |> 
  summarize(avg_mins = mean(day.mins), .by = churn)
  
```


What is the maximum length of international calls for loyal clients who do not use the international plan?

```{r}
df |> 
  filter(churn == "no", intl.plan == "no") |> 
  summarize(max_mins = max(intl.mins))
```

Creating some pivot tables for specific values of customer calls.

```{r}
df |> 
  select(area.code, day.calls, eve.calls, night.calls) |> 
  group_by(area.code) |> 
  summarize(across(everything(), mean))
```


Next, we are visually exploring the churn rate of clients with and without the international plan.

```{r}
df |> 
  ggplot(aes(x = intl.plan, fill = churn)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Customers having an international plan or not")
```

Users on the international plan are more likely to churn than those without the international plan.



Let us analyse the churn rate in relation to the number of customer calls to the call-center.


```{r}
df |> 
  mutate(customer.calls = factor(customer.calls)) |> 
  ggplot(aes(x = customer.calls, fill = churn)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Number of calls to customer service")
```

Observe that when the number of customer calls is larger then 3 the churn rate increases.



### Data wrangling

The dataset contains the follwing three categorical variables: `churn`, `voice.plan` and `intl.plan`. 

We convert all three variables to numerical variables having values of 0 (`no`) and 1 (`yes`).

```{r}
df <- df |> 
  mutate(
    churn = ifelse(df$churn == "yes", 1, 0),
    voice.plan = ifelse(df$voice.plan == "yes", 1, 0),
    intl.plan =  ifelse(df$intl.plan == "yes", 1, 0)
  )

```



We are considering two sets of features: the ones with the numerical attributes only (set "features") and only with additional categorical features "voice.plan","intl.plan" (set "features2")

```{r}
# choosing numerical features

features = c("voice.messages","intl.mins","intl.calls","intl.charge","day.mins", "day.calls", "day.charge","eve.mins","eve.calls", "eve.charge","night.mins","night.calls","night.charge","customer.calls","churn")

features2 = c("voice.plan","voice.messages","intl.plan","intl.mins","intl.calls","intl.charge","day.mins", "day.calls", "day.charge","eve.mins","eve.calls", "eve.charge","night.mins","night.calls","night.charge","customer.calls","churn")

for (i in features2){
  df[,i] = as.numeric(unlist(df[,i]))
}
```



::: {.callout-note icon=false}

The overall scheme of this experiment consists in applying several causal inference techniques, which give different types of results in various forms, and extract a consensus, if any, in the light of the different assumptions each model puts on the data.


:::


Performing the estimation of correlations between features.

```{r}
#| eval: false
dat = as.matrix(cor(df[,features2]))
rownames(dat) = NULL
colnames(dat) = NULL


## convert to tibble, add row identifier, and shape "long"
dat2 <-
  dat %>%
  as_tibble() %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "value") %>%
  mutate(
    Var1 = factor(Var1, levels = 1:17),
    Var2 = factor(gsub("V", "", Var2), levels = 1:17)
  )
#> Warning: The `x` argument of `as_tibble.matrix()` must have unique column names if
#> `.name_repair` is omitted as of tibble 2.0.0.
#> ℹ Using compatibility `.name_repair`.

ggplot(dat2, aes(Var1, Var2)) +
  geom_tile(aes(fill = value)) +
  geom_text(aes(label = round(value, 2)),size=1.75) +
  scale_fill_gradient(low = "blue", high = "red")+
  scale_x_discrete(name=c(""),breaks= 1:17,labels=features2) + 
  scale_y_discrete(name=c(""),breaks=1:17,labels =features2) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
ggsave("churn_corr.png")
```

# Specifying the train and test subsets of the data

```{r}
set.seed(2025)
df_split <- initial_split(df, prop = 2/3 )
df_train <- training(df_split)
df_test <- testing(df_split)
```

Prepare the data to apply bnlearn functions/prodecures



```{r}

Res<-pc.stable(df_train[,features2])

bnlearn:::print.bn(Res)

graph <- qgraph(Res, legend.cex = 0.35,
                asize=5,edge.color="black")
```

Model selection using the Markov blanket (Markov boundary) for the target variable $y = churn$ and using some information criteria. Using the logistic regression model to train to predict the probability of client to be churn.

```{r}

model_all = glm(churn ~., data = df_train[,features2], family = binomial(link = "logit"))

predictions = predict(model_all, newdata = df_test[,features2], type = "response")

acc = sum(round(predictions)==df_test$churn)/nrow(df_test)
sprintf("Accuracy of churn detection using all the variables %f", acc)
confusionMatrix(factor(round(predictions)), factor(df_test$churn), positive = as.character(1))
f_imp = coef(model_all)
print(f_imp)


model_sbset = glm(churn ~ voice.plan+intl.plan+intl.calls+customer.calls,data = df_train[,features2], family = binomial(link = "logit") )

predictions_sbset = predict(model_sbset, newdata = df_test[,features2], type = "response")
acc_sbst = sum(round(predictions_sbset)==df_test$churn)/nrow(df_test)
sprintf("Accuracy of churn detection using subset the variables %f", acc_sbst)


#### Prediction using the model selection procedure  

```

Visualize resulting feature importance in the logistic regression

```{r, fig.show="hold", out.width="20%"}
importance_data <- data.frame(Feature = names(f_imp), Importance = f_imp)

# Create a bar plot of feature importance
ggplot(importance_data, aes(x = Importance, y = Feature)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Feature Importance in Logistic Regression",
       x = "Importance",
       y = "Feature") +
  theme(axis.text.y = element_text(angle = 0, hjust = 1))  # Rotate x-axis labels for better readability
```

Different procedure to estimate the Markov blanket is given below. For example we consider the algorithm IAMB [2] and the Grow-Shrink (GS) [3,4] algorithm.

```{r}

Res_iamb<-iamb(df_train[,features2])

bnlearn:::print.bn(Res_iamb)

graph <- qgraph(Res_iamb, legend.cex = 0.35,
                asize=5,edge.color="black")


Res_gs<-gs(df_train[,features2])

bnlearn:::print.bn(Res_gs)

graph <- qgraph(Res_gs, legend.cex = 0.35,
                asize=5,edge.color="black")

```

***Refererences***

[1] Margaritis D (2003). Learning Bayesian Network Model Structure from Data. Ph.D. thesis, School of Computer Science, Carnegie-Mellon University, Pittsburgh, PA. Available as Technical Report CMU-CS-03-153.

[2] Spirtes, P., and Glymour, C., 1991. An Algorithm for Fast Recovery of Sparse Causal Graphs. Social Science Computer Review, vol. 9, Iss. 1, 62–72. <https://doi.org/10.1177/089443939100900106>

[3] P. Spirtes, C. Glymour, and R. Scheines. Causation, Prediction, and Search. MIT press, 2nd edition,2000. <https://doi.org/10.7551/mitpress/1754.001.0001>

[4] Tsamardinos, I., Aliferis, C.F. and Statnikov, A., 2003, August. Time and sample efficient discovery of Markov blankets and direct causal relations. In Proceedings of the ninth ACM SIGKDD international conference on Knowledge discovery and data mining (pp. 673-678). <https://doi.org/10.1145/956750.956838>

[5] Tsamardinos I, Aliferis CF, Statnikov A (2003). “Algorithms for Large Scale Markov Blanket Discovery.” In “Proceedings of the Sixteenth International Florida Artificial Intelligence Research Society Conference,” pp. 376–381. AAAI Press. In <https://doi.org/10.32473/flairs.v33i0.123621>
